---
title: "Southern Ocean CO2-system patterns and dynamics"
author: "Jens Daniel Müller"
date:  "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  workflowr::wflow_html:
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
editor_options:
  chunk_output_type: console
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

```{r library, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(scico)
```


```{r ggplot_theme, include = FALSE}
theme_set(theme_bw())
```

# Subsetting criteria

```{r subsetting_criteria}

SO_lat_high <- -30
SO_lat_low <- -65

year_start  <- 2004
year_end    <- 2017
depth_lower_limit <- 500

f <- 2
qc <- 1

```

Southern Ocean is defined here as the area between:

- higher latitude: `r SO_lat_high`
- lower latitude:  `r SO_lat_low`

Time is restricted to years from:

- `r year_start` to `r year_end`

Only observations with:

- maximum depth: `r depth_lower_limit`
- quality flag f: `r f`
- quality flag qc: `r qc`

were investigated. Quality flags were checked individually for `tco2` and `talk`. The resulting data sets were merged, such that each accepted `tco2` and `talk` is kept and we have observations with only one of both parameters determined.

```{r subset_GLODAP, eval=FALSE}

GLODAP <- read_csv(here::here("data/_summarized_data_files",
                              "GLODAPv2.2020_Merged_Master_File_JM.csv"),
                   guess_max = 1e5)

# GLODAP_SO <- GLODAP_SO %>% 
#   mutate(decade = as.factor(decade))
GLODAP_SO <- GLODAP %>% 
  filter(latitude <= SO_lat_high, latitude >= SO_lat_low)

GLODAP_SO <- GLODAP_SO %>% 
  filter(year >= year_start, year <= year_end)

GLODAP_SO_tco2 <- GLODAP_SO %>% 
  filter(tco2f == f,
         tco2qc == qc,
         !is.na(tco2))

GLODAP_SO_talk <- GLODAP_SO %>% 
  filter(talkf == f,
         talkqc == qc,
         !is.na(talk))

GLODAP_SO <- full_join(GLODAP_SO_tco2, GLODAP_SO_talk)
rm(GLODAP_SO_talk, GLODAP_SO_tco2)

GLODAP_SO <- GLODAP_SO %>% 
  filter(depth <= depth_lower_limit)


GLODAP_SO <- GLODAP_SO %>% 
  select(date, cruise:bottomdepth, depth:salinity, tco2, talk)


GLODAP_SO %>% write_csv(here::here("data/_summarized_data_files",
                                   "GLODAPv2.2020_Merged_Master_File_JM_SO.csv"))

```

```{r reopen_subsetted_GLODAP}

rm(depth_lower_limit, f, qc, SO_lat_high, SO_lat_low, year_end, year_start)

GLODAP_SO <- read_csv(here::here("data/_summarized_data_files", "GLODAPv2.2020_Merged_Master_File_JM_SO.csv"),
                   guess_max = 1e5)

GLODAP_SO_long <- GLODAP_SO %>% 
  pivot_longer(tco2:talk, names_to = "var", values_to = "value")


```

# Gridding

## Vertical

```{r depth_levels}

depth_levels <-
  c(2.5, 
    5, 
    seq(10,50,10), 
    seq(75,150,25), 
    seq(200,500,50), 
    seq(600,1900,100), 1975)

depth_intervals <-
  c(replace_na(depth_levels - (depth_levels - lag(depth_levels))/2, 0),
    2000)

```

As an adaptation of Lydia's interpolation to 33 depth levels:

`r depth_levels`

data were clustered into depth intervals between following depth limits:

`r depth_intervals`

However, the intervals were labeled with the exact same numbers as used by Lydia. This grouping in comparison to linear interpolation avoids creating "unobserved" values at the cost of less precise vertical information.

```{r vertical_gridding}

GLODAP_SO_long <- GLODAP_SO_long %>% 
  mutate(depth_grid = as.numeric(as.character(cut(depth,
                          breaks = depth_intervals,
                          labels = depth_levels,
                          include.lowest = TRUE))))

GLODAP_SO <- GLODAP_SO %>% 
  mutate(depth_grid = as.numeric(as.character(cut(depth,
                          breaks = depth_intervals,
                          labels = depth_levels,
                          include.lowest = TRUE))))

# GLODAP_SO_long_depth_grid <- GLODAP_SO_long %>% 
#   select(date, year, month, latitude, longitude, tco2, talk, depth, depth_grid) %>% 
#   group_by(depth_grid, year, month) %>% 
#   summarise_all(lst(mean, sd, min, max), na.rm=TRUE) %>% 
#   ungroup()

```

## Horizontal

Observations were gridded to a 1° x 1° degree latitude and longitude grid.

For exploratory analysis, data were also clustered into coarse 20°-longitude and 5°-latitude grids.

```{r horizontal_gridding}

GLODAP_SO_long <- GLODAP_SO_long %>% 
  mutate(latitude_grid = as.numeric(as.character(cut(latitude,
                                                     breaks =  seq(-90, 90, 1),
                                                     labels = seq(-89.5,89.5,1)))),
         longitude_grid = as.numeric(as.character(cut(longitude,
                                                      breaks =  seq(-180, 180, 1),
                                                      labels = seq(-179.5,179.5,1)))),
         latitude_grid_coarse = cut(latitude, seq(-90,90,5)),
         longitude_grid_coarse = cut(longitude, seq(-180, 180, 40)))

```


# Spatial distribution

## Surface map

All observations (`tco2` and/or `talk`), were plotted with each sampling year being displayed as color or indivdual facet. Sampling depth was not taken into account at his stage, but it is assumed that mainly completed profiles were submitted to GLODAP.

It appears that:

- homogenious distribution of observations in space and time is questionable, which can be critical for trend and variability analysis
- Most sampling events are organized either meridional or zonal
- in 2009/10 many observations exist in 30-35°S 

```{r observations_map}

map_southpole <- borders("world", colour="gray60", fill="gray60", ylim = c(-90,-25)) # create a layer of borders

GLODAP_SO_long %>% 
  ggplot(aes(longitude, latitude, col=as.factor(year))) +
  map_southpole+
  geom_hline(yintercept = seq(-65, -30, 5)) +
  geom_vline(xintercept = seq(-180, 180, 40)) +
  geom_point() +
  coord_map("ortho", orientation = c(-90, 0, 0),
            ylim = c(-85,-30), xlim = c(-180,180)) +
  scale_color_scico_d(palette = "bamako", name="year") +
  #scale_color_viridis_d(option = "magma", name="year") +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  facet_wrap(~var) +
  theme(legend.position = "bottom")
  
```

```{r observations_map_year_grid, fig.asp=1.5}

GLODAP_SO_long %>% 
  ggplot(aes(longitude, latitude)) +
  map_southpole+
  geom_hline(yintercept = seq(-65, -30, 5)) +
  geom_vline(xintercept = seq(-180, 180, 40)) +
  geom_point() +
  coord_map("ortho", orientation = c(-90, 0, 0),
            ylim = c(-85,-30), xlim = c(-180,180)) +
  scale_color_viridis_d(name="year") +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  facet_wrap(~year, ncol = 3)
  
```

# Time series

## Observed values

Observed `tco2` and `talk` for selected depth levels were plotted, colored by coarse latitute and longitude grids.

First impressions are:

- `tco2` is lower than `talk` and more variable (as expected)
- `tco2` increases with latitude
- latitudinal patterns less pronounced for `talk`
- longitudinal patterns less pronounced for both parameters
- dip around 2009 might be biased by sampling pattern

```{r tCO2_time_series_lon_grid, fig.asp=1}

GLODAP_SO_long %>% 
  filter(depth_grid %in% c(10, 100, 300, 500)) %>% 
  ggplot(aes(date, value, col=longitude_grid_coarse)) +
  geom_vline(xintercept = ymd("2009-01-01")) +
  geom_point() +
  facet_grid(depth_grid~var) +
  theme(axis.title.x = element_blank()) +
  theme(legend.position = "bottom")

```

```{r tCO2_time_series_lat_grid, fig.asp=1}

GLODAP_SO_long %>% 
  filter(depth_grid %in% c(10, 100, 300, 500)) %>% 
  ggplot(aes(date, value, col=latitude_grid_coarse)) +
  geom_vline(xintercept = ymd("2009-01-01")) +
  geom_point() +
  scale_color_viridis_d() +
  facet_grid(depth_grid~var) +
  theme(axis.title.x = element_blank()) +
  theme(legend.position = "bottom")

```

## Mean values

### Zonal

In order to minimize the effect of latitudinal sampling bias in the time series plots, the annual mean `tco2` and `talk` were calculted for 5°-latitude intervals.

First impressions:

- 2009 `tco2` dip only detectable in 500m and intermediate latitudes
- similar dip seems to be detectable in `talk`, questioning the MOC hypothesis

```{r calculate_annual_mean_value_lat_grid}

GLODAP_SO_long_mean_lat <- GLODAP_SO_long %>% 
  group_by(year, latitude_grid_coarse, depth_grid, var) %>% 
  summarise(value_mean = mean(value, na.rm=TRUE)) %>% 
  ungroup() %>% 
  arrange(depth_grid, var, year) %>% 
  filter(!is.na(value_mean))


```

```{r time_series_annual_mean_lat_grid, fig.asp=1}

GLODAP_SO_long_mean_lat %>% 
  filter(depth_grid %in% c(10, 100, 300, 500)) %>% 
  ggplot(aes(year, value_mean, col=latitude_grid_coarse)) +
  geom_vline(xintercept = 2009) +
  geom_line() +
  geom_point() +
  scale_color_viridis_d() +
  facet_grid(depth_grid~var) +
  theme(axis.title.x = element_blank()) +
  theme(legend.position = "bottom")

```

### Entire SO

In order to proof the effect of latitudinal sampling bias in the time series plots, the annual mean `tco2` and `talk` were calculted for the entire SO.

First impressions:

- 2009 `tco2` dip only detectable in 500m and intermediate latitudes
- similar dip seems to be detectable in `talk`, questioning the MOC hypothesis

```{r calculate_annual_mean_value}

GLODAP_SO_long_mean <- GLODAP_SO_long %>% 
  group_by(year, depth_grid, var) %>% 
  summarise(value_mean = mean(value, na.rm=TRUE)) %>% 
  ungroup() %>% 
  arrange(depth_grid, var, year) %>% 
  filter(!is.na(value_mean))


```

```{r time_series_annual_mean}

GLODAP_SO_long_mean %>% 
  filter(depth_grid %in% c(10, 100, 300, 500)) %>% 
  ggplot(aes(year, value_mean)) +
  geom_vline(xintercept = 2009) +
  geom_line(aes(col=as.factor(depth_grid))) +
  geom_point(aes(fill=as.factor(depth_grid)),shape=21) +
  scale_fill_scico_d(palette = "batlow", name="Depth levels (m)", direction = -1) +
  scale_color_scico_d(palette = "batlow", name="Depth levels (m)", direction = -1) +
  facet_wrap(~var) +
  theme(axis.title.x = element_blank()) +
  theme(legend.position = "bottom")

```

## Number of observations

### Zonal

The annual number of observation were calculted for 5 degree latitude intervals.

- Outstanding high number of observations in 2009 at 30-35°S


```{r count_observations_lat_grid}

GLODAP_SO_long_nr_lat <- GLODAP_SO_long %>% 
  filter(!is.na(value)) %>% 
  group_by(year, latitude_grid_coarse, depth_grid, var) %>% 
  summarise(count = n()) %>% 
  ungroup()

```


```{r time_series_lat_grid_counts, fig.asp=1}

GLODAP_SO_long_nr_lat %>% 
  filter(depth_grid %in% c(10, 100, 300, 500)) %>% 
  ggplot(aes(year, count, col=latitude_grid_coarse)) +
  geom_vline(xintercept = 2009) +
  geom_line() +
  geom_point() +
  scale_color_viridis_d() +
  facet_grid(depth_grid~var) +
  theme(axis.title.x = element_blank()) +
  theme(legend.position = "bottom")

```


# Spatial patterns

In order to order to gain a more precise idea about the spatial distribution of `tco2` and `talk`, both parameters were plotted against latitude and colored by longitude, and vice versa.

```{r latitudinal_patterns, fig.asp=1}

GLODAP_SO_long %>% 
  filter(depth_grid %in% c(10, 100, 300, 500)) %>% 
  ggplot(aes(latitude, value, col=longitude_grid_coarse)) +
  geom_point() +
  facet_grid(depth_grid~var) +
  theme(legend.position = "bottom")

```

```{r longitudinal_patterns, fig.asp=1}

GLODAP_SO_long %>% 
  filter(depth_grid %in% c(10, 100, 300, 500)) %>% 
  ggplot(aes(longitude, value, col=latitude_grid_coarse)) +
  geom_point() +
  scale_color_viridis_d() +
  facet_grid(depth_grid~var) +
  theme(legend.position = "bottom")

```

# Correlations

## tco2/talk vs temperature

`tco2` and `talk` were plotted against temperature.

First impression:
- DIC depends more strongly on temperature, likely due to the higher solubilty of cold waters

```{r tco2_talk_vs_temp_correlation, fig.asp=1}

GLODAP_SO_long %>% 
  filter(depth_grid %in% c(10, 100, 300, 500)) %>% 
  ggplot(aes(temperature, value, col=latitude)) +
  geom_point() +
  scale_color_viridis_c() +
  facet_grid(depth_grid~var) +
  theme(legend.position = "bottom")


```

## tco2 vs talk

```{r tco2_vs_talk_correlation}

GLODAP_SO %>% 
  filter(depth_grid %in% c(10, 100, 300, 500)) %>% 
  ggplot(aes(talk, tco2, col=latitude)) +
  geom_point() +
  scale_color_viridis_c() +
  facet_wrap(~depth_grid)

```


# Open tasks / questions

- Check duplicated analysis at identical sampling location and time
- transect plot for 2009 meridional cruise
- plot worldmap and time series of GLODAP observation counts 2004-17
- time series of zonal annual mean values referring to Orsi-fronts

# Checks with Lydia

- number of tco2 observations in SO


